---
title: "RepProj1"
---

### Read the data file
```{r}
#unzip the file
unzip("activity.zip")

#Read the data file into R
Act<-read.csv("activity.csv")
```

### Mean number of steps per day
1. Calculate the total steps per day
2. Make a Histogram
3. Calculate the mean and total number of steps per day
```{r}
steps<-tapply(Act$steps,Act$date,FUN=sum)
hist(steps)
```

The mean number of steps per day is `r mean(steps,na.rm=TRUE)`.  and the median is `r median(steps,na.rm=TRUE)`.

### Average daily activity pattern
```{r}
intMean<-tapply(Act$steps,Act$interval,FUN=mean,na.rm=TRUE)
plot(names(intMean),intMean,type="l",xlab="Interval",ylab="Mean Steps")
names(which.max(intMean))
```

This is the 5-minute interval with the maximum average number of steps. It is given in the form HHMM, i.e. hour and minute of the day, so 835 is the 5 minute interval from 8:35 AM to 8:40 AM.

### Imputing missing values

1. Count the number of rows in the datafram with na's 
```{r}
sum(is.na(Act))
```

2. Strategy to replace the na's in the dataframe: use values that are the mean of that 5 minute interval across all days.

3. Create the new dataframe.

```{r}
intcol<-names(intMean) # extract intervals
temp<-cbind(intMean,intcol) # combine intervals and mean values

newsteps=merge(Act,temp,by.x="interval",by.y="intcol",all=TRUE) # add mean values column to dataframe

# Replace na's with interval means
for(i in 1:length(Act$steps)) {
  if(is.na(newsteps$steps[i]))
    {newsteps$steps[i]<-as.integer(newsteps$intMean[i])}
}
newsteps$intMean<-NULL # Don't need this column anymore
                           
```

4. Histogram of total number of steps per day.

```{r}
steps<-tapply(newsteps$steps,newsteps$date,FUN=sum)
hist(steps)
```

- Calculate the mean and median:  
 
The mean number of steps per day is `r mean(steps,na.rm=TRUE)`.  and the median is `r median(steps,na.rm=TRUE)`.  

- Do these values differ from the previous estimates?  

Yes, they are somewhat larger than before.  

- What is the impact of imputing missing data on the estimates of the total daily number of steps?  

The data are now in two groups, a group that is roughly normally distributed, and a group of outliers where the number of steps per day is much higher.

### Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable with levels "weekday" and "weekend".
2. Make a panel plot with a time series plot (type="l") of the 5 minute interval (x-axis) and average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}
# add a column, days of the week
newsteps$day<-as.character(weekdays(as.Date(newsteps$date)))

# convert to weekday/weekend
for(i in 1:length(newsteps$steps)) {
  if(newsteps$day[i]=="Saturday"|newsteps$day[i]=="Sunday"){
    newsteps$day[i]<-"weekend"} 
  else{
    newsteps$day[i]<-"weekday"
    }
}
newsteps$day<-as.factor(newsteps$day) # change to a factor variable

library(lattice)
xyplot(steps~interval|day,data=newsteps,type="l")
```