---
title: "Work Done by an Expanding Gas"
author: "NeverZero"
date: "Monday, February 09, 2015"
output: html_document
---

The calculations below relate to an experiment with parameters described in reference [1].

The heat of combustion of $H_2$ is 286 KJ/mole [2]. For 1.04E-3 mole, that is an energy of `r 286E3*1.04E-3` J.

Work done by an expanding gas on its environment in an isothermal (constant temperature) process [3]:

$$W=-\int_{V_a}^{V_b}\mathrm{P}\,\mathrm{d}V=-\int_{V_a}^{V_b}\frac{nRT}{V}\,\mathrm{d}V=-nRT*ln(\frac{V_b}{V_a})$$

(Negative work means the environment has done positive work on the system.)

```{r}
n<-1.04E-3 # moles of H2
R<-8.314 # J/K-mol
T<-1300 # K
Pa=1500/14.5*1E5 # Pa (Note 1 bar=14.5 psi, 1 bar=1e5 Pa) 
Pb= 1e5 # ~Atmospheric pressure in Pa
#Vb<-n*22.4*(1000)/(100^3)*(1300/300) # m^3, at same T
Va<-1.06e-6 # m^3
Vb=Pa*Va/Pb
W<-(-n*R*T)*log(Vb/Va)
W
```


The work done by an adiabatic expansion process [4]:

$$W = \int_{V_1}^{V_2}P_1 \left(\frac{V_1}{V} \right)^\gamma\, dV$$

```{r}
gamma=1.358 # For H2 at 1000 C, per wiki
alpha=3 # 1/(gamma-1) # Degrees of freedom/2 for H2
Vb= n*22.4*1000/100^3
W=-alpha*Pa*Va*((Vb/Va)^(1-gamma)-1)
W
```


Adiabatic expansion, second method [5]:

$$W = (P_a*V_a - P_b*V_2)/(\gamma - 1)$$

```{r}
W=(Pa*Va-Pb*Vb)/(gamma-1)
W
```

Compare this to the approximately 460 J available in 50 mg of flashpowder, the maximum allowed in a firecracker in the US [6]. (The energy density of flash powder is ~2.3 Kcal/g [7,8].) 

Recalculate at 20 bar

```{r}
Pa=20*1e5
W=(Pa*Va-Pb*Vb)/(gamma-1)
W
```

1. [Bang!](https://docs.google.com/document/d/1QHxqzC5umWBdvNOMQTvoi7pdxiZ8h54tE_KX2a_41pE/edit)  
2. [Heat of Combustion](http://en.wikipedia.org/wiki/Heat_of_combustion)  
3. [Isothermal Process](http://en.wikipedia.org/wiki/Isothermal_process)  
4. [Adiabatic Process](http://en.wikipedia.org/wiki/Adiabatic_process)  
5. [Thermal Processes](http://faculty.wwu.edu/vawter/physicsnet/Topics/Thermal/ImportantThermalProcess.html)  
6. [M-80 Firecracker](http://en.wikipedia.org/wiki/M-80_%28explosive%29)  
7. [Flash Powder Energy Density](http://www.powerlabs.org/chemlabs/deflagrants.htm)  
8. [Their Arrows will Darken the Sun by Mark Denny](https://books.google.com/books?id=pxScgdDCQVUC&lpg=PA173&ots=EwtDzFg5UC&dq=heat%20of%20combustion%20of%20flash%20powder&pg=PA173#v=onepage&q=heat%20of%20combustion%20of%20flash%20powder&f=false)  
